Types from Data: Making Structured Data First-Class Citizens in F\#

### 2016-05-03 pldi-2016/paper.tex
Required publication changes for PLDI 2016

\documentclass[10pt]{sigplanconf}

\usepackage{listings}
\usepackage{enumitem}
\usepackage[hyphens]{url}
\usepackage[svgnames]{xcolor}
\definecolor{lcl}{RGB}{140,0,100}
\usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
\hypersetup{urlcolor=lcl,linkcolor=lcl,citecolor=lcl}
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics}
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{ftnright}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage[hang,flushmargin]{footmisc}
\usepackage[notquote]{hanging}
\usepackage{flushend}
\usepackage{enumerate}
\usepackage{etoolbox}

@@

@@

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}
\conferenceinfo{PLDI '16}{June 13--17, 2016, Santa Barbara, CA, United States}
\copyrightyear{2016}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
%\doi{nnnnnnn.nnnnnnn}

@@

\title{Types from data: \textnormal{Making structured data first-class citizens in F\#}}
%\subtitle{Subtitle Text, if any}

\authorinfo{Tomas Petricek}
           {University of Cambridge}
           {tomas@tomasp.net}
\authorinfo{Gustavo Guerra}
           {Microsoft Corporation, London}
           {gustavo@codebeside.org}
\authorinfo{Don Syme}
           {Microsoft Research, Cambridge}
           {dsyme@microsoft.com}
\maketitle

@@

\category{D.3.3}{Programming Languages}{Language Constructs and Features }
\keywords F\#, Type Providers, Inference, JSON, XML

@@

For example, \url{http://openweathermap.org/current} contains one example to document an end-point
to get the current weather. Using standard libraries, we might call it as\footnote{We abbreviate
the full URL and omit application key (available after registration). The returned JSON is shown
in Appendix~\ref{sec:appendix-weather} and can be used to run the code against a local file.}:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format}
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \strf{main} field, which has to be another record containing
a numerical \strf{temp} field representing the current temperature. When the shape is different,
the code fails. While not immediately unsound, the code is prone to errors if strings are
misspelled or incorrect shape assumed.

Using the JSON type provider from F\# Data, we can write code with exactly the
same functionality in two lines:
%
\vspace{-0.1em}
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Lovely \%f!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider \cite{fsharp-typeprov} at
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

@@

\section{Type providers for structured data}
\label{sec:providers}

@@

\subsection{Working with JSON documents}
\label{sec:providers-json}

@@

\subsection{Processing XML documents}
\label{sec:providers-xml}

@@

\subsection{Real-world JSON services}
\label{sec:providers-sum}

@@

\section{Shape inference for structured data}
\label{sec:inference}

@@

\subsection{Inferred shapes}
\label{sec:inference-types}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Important aspects of the preferred shape relation}
\label{fig:subtyping-diagram}
\end{figure}

@@

\subsection{Preferred shape relation}
\label{sec:inference-subtyping}

@@

\subsection{Common preferred shape relation}
\label{sec:inference-commonsuper}

@@

\subsection{Inferring shapes from samples}
\label{sec:formal-inferval}

@@

\begin{figure}[!h]
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
 \\[0.5em]
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.6em]
 \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsubseteq\ \textnormal{lifted over substitutions} \\[0.1em]
\end{array}
\end{equation*}
\caption{Shape inference from sample data}
\label{fig:shape-inference}
\vspace{-1em}
\end{figure}

@@

\subsection{Adding labelled top shapes}
\label{sec:inference-vars}

@@

\section{Formalizing type providers}
\label{sec:formal}

@@

@@

\subsection{The Foo calculus}
\label{sec:formal-ff}

@@

\subsection{Type providers}
\label{sec:formal-tp}

@@

\section{Relative type safety}
\label{sec:safety}

@@

\section{Practical experience}
\label{sec:impl}

@@

\subsection{Relative safety in practice}
\label{sec:safety-discuss}

@@

\subsection{Parsing structured data}
\label{sec:impl-parsing}

@@

\subsection{Providing idiomatic F\# types}
\label{sec:impl-naming}

@@

\subsection{Heterogeneous collections}
\label{sec:impl-hetero}

@@

\subsection{Predictability and stability}
\label{sec:impl-stable}

@@

\section{Related and future work}
\label{sec:related}

>>>>

\documentclass[pldi-cameraready]{sigplanconf-pldi16}

\RequirePackage[loading]{tracefnt}

\usepackage{listings}
\usepackage{enumitem}
\usepackage[hyphens]{url}
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{graphics}
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{ftnright}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage[hang,flushmargin]{footmisc}
\usepackage[notquote]{hanging}
\usepackage{flushend}
\usepackage{enumerate}
\usepackage{etoolbox}

@@

\CopyrightYear{2016}
%\setcopyright{acmlicensed}
\conferenceinfo{PLDI '16,}{June 13 - 17, 2016, Santa Barbara, CA, USA}
\copyrightdata{978-1-4503-4261-2/16/06}
%\acmPrice{\$15.00}
\copyrightdoi{2908080.2908115}

@@

@@

\title{Types from Data: Making Structured Data First-Class Citizens in F\#}
%\subtitle{Subtitle Text, if any}

\authorinfo{Tomas Petricek}
           {University of Cambridge, UK}
           {tomas@tomasp.net}
\authorinfo{Gustavo Guerra}
           {Microsoft Corporation, UK}
           {gustavo@codebeside.org}
\authorinfo{Don Syme}
           {Microsoft Research Cambridge, UK}
           {dsyme@microsoft.com}
\maketitle

@@

\category{D.3.3}{Programming Languages}{Language Constructs and Features }
\terms{Languages, Theory}
\keywords F\#, Type Providers, Inference, JSON, XML

@@

For example, \url{http://openweathermap.org/current} contains one example to document an end-point
to get the current weather. Using standard libraries, we might call it as\footnote{We abbreviate
the full URL and omit application key (available after registration). The returned JSON is shown
in Appendix~\ref{sec:appendix-weather} and can be used to run the code against a local file.}:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format}
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \strf{main} field, which has to be another record containing
a numerical \strf{temp} field representing the current temperature. When the shape is different,
the code fails. While not immediately unsound, the code is prone to errors if strings are
misspelled or incorrect shape assumed.

Using the JSON type provider from F\# Data, we can write code with exactly the
same functionality in two lines:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Lovely \%f!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider \cite{fsharp-typeprov} at
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

@@

\section{Type Providers for Structured Data}
\label{sec:providers}

@@

\subsection{Working with JSON Documents}
\label{sec:providers-json}

@@

\subsection{Processing XML Documents}
\label{sec:providers-xml}

@@

\subsection{Real-world JSON Services}
\label{sec:providers-sum}

@@

\section{Shape Inference for Structured Data}
\label{sec:inference}

@@

\subsection{Inferred Shapes}
\label{sec:inference-types}

@@

\begin{figure}
\begin{center}
% \includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\includegraphics[scale=0.80,clip]{images/hierarchy.png} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Important aspects of the preferred shape relation}
\label{fig:subtyping-diagram}
\end{figure}

@@

\subsection{Preferred Shape Relation}
\label{sec:inference-subtyping}

@@

\subsection{Common Preferred Shape Relation}
\label{sec:inference-commonsuper}

@@

\subsection{Inferring Shapes from Samples}
\label{sec:formal-inferval}

@@

@@

\subsection{Adding Labelled Top Shapes}
\label{sec:inference-vars}

@@

\section{Formalizing Type Providers}
\label{sec:formal}

@@

\begin{figure}[!h]
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\begin{equation*}
\begin{array}{l}
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
 \\[0.5em]
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.6em]
 \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsubseteq\ \textnormal{lifted over substitutions} \\[0.1em]
\end{array}
\end{equation*}
\caption{Shape inference from sample data}
\label{fig:shape-inference}
\end{figure}

@@

\subsection{The Foo Calculus}
\label{sec:formal-ff}

@@

\subsection{Type Providers}
\label{sec:formal-tp}

@@

\section{Relative Type Safety}
\label{sec:safety}

@@

\section{Practical Experience}
\label{sec:impl}

@@

\subsection{Relative Safety in Practice}
\label{sec:safety-discuss}

@@

\subsection{Parsing Structured Data}
\label{sec:impl-parsing}

@@

\subsection{Providing Idiomatic F\# Types}
\label{sec:impl-naming}

@@

\subsection{Heterogeneous Collections}
\label{sec:impl-hetero}

@@

\subsection{Predictability and Stability}
\label{sec:impl-stable}

@@

\section{Related and Future Work}
\label{sec:related}
